---
title: "R Code for Binomial based Tests"
format: html
editor: visual
self-contained: true
---

## The binomial Test -p- Example from Lecture Notes

#### Example 1:

- From studies 50% of men suffer from side effects of a surgery.
- New method of performing the surgery is tested.
- Out of 19 operations only 3 men suffered from the side effects.

Question: is it safe to conclude that the new method is effective?


To answer the question we run a lower tailed binomial test.

```{r}
binom.test(x=3, # #successes
          n=19, # total number of operations
          p= 0.5, # hypothesized probability
          alternative="l" # lower tailed test
          )
```

#### Example 2:

To answer the Mendelian example from the notes we run a two tailed binomial test.

```{r}
binom.test(x=682, # #tall plants
          n=925, # total number of plants
          p= 0.75, # hypothesized probability
          alternative="t" # two tailed test
          )
```



## The quantile Test -x_p- Example from Lecture Notes

#### Example 3:
Assume X is the score of a high school examination to go to college. Previous studies showed that the upper quartile of these scores is $X_{0.75} = $ 193.

15 graduate high school students scores are given:

189, 233, 195, 160, 212, 176, 231, 185, 199, 213, 202, 193, 174, 166, 248.

To test whether the population upper quartile is still 193, we run a two-tailed quantile test.

```{r}
# data
x= c(189, 233, 195, 160, 212, 176, 231, 185, 199, 213, 202, 193, 174, 166, 248)

# Test statistics
T1 = sum(x<=193)
T2 = sum(x<193)

# p-value
p = 2*min(pbinom(T1,15,0.75),pbinom(T2-1,15,0.75,lower.tail = F))
p
```



## The sign Test - Example from Lecture Notes

#### Example 4:
- Item A (old process) and Item B (new process)
- Out of 10 consumers:
  -   8 preferred B over A ==> +
  -   1 preferred A over B ==> -
  -   1 had no preference ==> 0
- Do consumers prefer B over A?
- To answer this we run an upper tailed sign test.

Th sign test is a binomial test with p=0.5.

```{r}
binom.test(x=8,
           n=9,
           p=0.5,
           alternative = "g")
```




#### Example 5:
- 6 students went on a diet to lose weight. A Pair is (WB, WA)
  -   5 lost weight (WB > WA) ==> -
  -   1 gained weight (WB < WA) ==> +
- Is the diet effective?
- To answer this we run an lower tailed sign test.

Th sign test is a binomial test with p=0.5.

```{r}
binom.test(x=1,
           n=6,
           p=0.5,
           alternative = "l")
```



## Tolerance Limits

#### Example 6:
- Electric seat adjusters.
- A car manufacture wants to know how much range of adjustment is needed to be 90% sure that at least 80% of the population of buyers will be able to adjust their seats.

```{r}
#install.packages("tolerance") # install package
library(tolerance)

# How to find Q
distfree.est(n = 18, # sample size
             alpha = 0.1, # 1-alpha is the confidence
             P= NULL,
             side = 2. # two sided interval
               )

# How to find n
distfree.est(n = NULL, # sample size
             alpha = 0.1, # 1-alpha is the confidence
             P= 0.8,
             side = 2. # two sided interval
               )
```

