---
title: "Contingency Tables"
format: html
author: Dr. Cohen 
---

## Contingency Tables

`Definition` A contingency table is an array of natural numbers in matrix form where those numbers represent counts / frequencies.


|         | Col 1 | Col2 | Totals
|---------|:-----:|:------:|:------:|
| **row 1**   | $O_{11}$   |    $O_{12}$  |   $r_{1}$    |
| **row 2**   | $O_{21}$      |   $O_{22}$  |  $r_{2}$    |
| **Totals**  | $c_{1}$   |   $c_{2}$  |  N   |

 2 x 2 contingency table


### Chi-squared Test for differences in Probabilities

`Data` 


|         | Class 1 | Class2 | `Totals`
|---------|:-----:|:------:|:------:|
| **Population 1**   | $O_{11}$ ($p_1$)   |    $O_{12}$ |   $n_1$   |
| **Population 2**   | $O_{21}$  ($p_2$)   |   $O_{22}$ |  $n_2$   |
| **`Totals`**  | $c_1$  |   $c_2$ | N    |

 2 x 2 contingency table
 
 
 `Assumption` 
 
- Each sample is Random Sample
- The 2 samples are independent
- Each observation can be classified into class 1 and class 2


`Test Statistic`

$$ T= \frac{\sqrt{N} (O_{12}O_{22} - O_{12}O_{21})}{\sqrt{n_1n_2c_1c_1}}$$
`Null distirbution`: $T \sim N(0,1)$


`Hypothesis: Two-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 \neq p_2 $$
- $p_1$ the probability that a randomly selected obs from the population 1 will be in class 1.
- $p_2$ the probability that a randomly selected obs from the population 2 will be in class 1.

- `P-value`$= 2\times \min\{  P( T \leq T_{Obs}), P(T \geq T_{Obs}) \}$

- `Decision`: If p_value < $\alpha$ then REJECT $H_0$


`Hypothesis: Lower-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 < p_2 $$
- $p_1$ the probability that a randomly selected obs from the population 1 will be in class 1.

- `P-value`$= P( T \leq T_{Obs})$

- `Decision`: If p_value < $\alpha$ then REJECT $H_0$


`Hypothesis: Upper-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 > p_2 $$
- $p_1$ the probability that a randomly selected obs from the population 1 will be in class 1.

- `P-value`$= P( T \geq T_{Obs})$

- `Decision`: If p_value < $\alpha$ then REJECT $H_0$

## Chi-squared Test - Example 1

The number of items in two car loads.

`Data` 


|         | Defective | Non defective | `Totals`
|---------|:-----:|:------:|:------:|
| **Carload 1**   | a =13   |    b=73 |   $n_1$ = 86   |
| **Carload 2**   | c  = 17   |   d=57 |  $n_2$ = 74   |
| **`Totals`**  | $c_1$ = 30 |   $c_2$= 130 | N = 160  |

 2 x 2 contingency table
 
 `Question`: Test whether there are differences in proportions of defective items between the two carloads.
 
1. Define the null and alternative hypotheses <br>
`Answer:` $$ H_0: p_1 = p_2 $$ $$H_1: p_1 \neq p_2 $$
This is an two-tailed test.
2. Find the Test statistic observed and null distribution<br>
`Answer:` $T_{obs}=-1.2695$ and $T\sim N(0,1)$
3. Determine critical values (rejection region) <br>
`Answer:` +/- `r {qnorm(0.025)}`
4. Find P-value<br>
`Answer:` p-value=`r {2*pnorm(-1.2695)}`
5. Decision
<br>
`Answer:` Since P-value > 0.05 then Fail to Reject $H_0$.



## Chi-squared Test - Example 1 with R

```{r, echo=TRUE}
data = cbind(c(13,17),c(73,57)) # create data
chisq.test(data, # table data
           correct = FALSE # find p-value without Yates' correction
           ) 
chisq.test(data, # table data
           correct = TRUE # find p-value with Yates' correction
           ) 
```

- `Interpretation`: Fail to Reject $H_0$. There is evidence to support that the data is compatible with equal proportions $p-value=0.2$. 
- Note: $T_{obs}^2 = (-1.2695)^2$ = `r {(-1.2695)^2}` 



## Chi-squared Test - Example 2

A new toothpaste is tested for men and women preferences.

`Data` 


|         | Like | Do not like | `Totals`
|---------|:-----:|:------:|:------:|
| **Men**   | a =64   |    b=36 |   $n_1$ = 100   |
| **Women**   | c  = 74   |   d=26 |  $n_2$ = 100   |
| **`Totals`**  | $c_1$ = 138 |   $c_2$= 62 | N = 200  |

 2 x 2 contingency table
 
 `Question`: Do men and women differ in their preferences regarding the new toothpaste?
 

1. Define the null and alternative hypotheses <br>
`Answer:` $$ H_0: p_1 = p_2 $$ $$H_1: p_1 \neq p_2 $$
This is an two-tailed test.
2. Find the Test statistic observed and null distribution<br>
`Answer:` $T_{obs}=-1.53$ and $T\sim N(0,1)$
3. Determine critical values (rejection region) <br>
`Answer:` +/- `r {qnorm(0.025)}`
4. Find P-value<br>
`Answer:` p-value=`r {2*pnorm(-1.53)}`
5. Decision and Interpretation
<br>
`Answer:` Since P-value > 0.05 then Fail to Reject $H_0$.

There is insufficient evidence to support that men and women differ in their preferences regarding the new toothpaste. 






 
## Fisher's Exact Test

`Data` 


|         | col 1 | col 2 | `Totals`
|---------|:-----:|:------:|:------:|
| **row 1**   | X ($p_1$)   |    r-X |   r   |
| **row 2**   | c-X  ($p_2$)   |   N-r-c+X|  N-r  |
| **`Totals`**  | c |   N-c | N    |

 2 x 2 contingency table
 
 
 `Assumption` 
 
- Each observation can be in one cell
- The row and column totals are fixed.


## Fisher's Exact Test

`Test Statistic`

T = X = number of obs. in row 1 and col 1.

$$ T (H_0) \sim hypergeometric(N,r,C) $$
The PMF is:

$$ P(T=x) = \frac{\binom{r}{x}\binom{N-r}{c-x}}{\binom{N}{c}} $$
x=0,1,2,...,min(r,m)



## Fisher's Exact Test

`Hypothesis: Two-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 \neq p_2 $$
- $p_1$ the probability that a randomly selected obs from the row 1 will be in col 1.

- `P-value`$= 2\times \min\{  P( T \leq T_{Obs}), P(T \geq T_{Obs}) \}$

- `Decision`: IF p_value < $\alpha$ then REJECT $H_0$



### Fisher's Exact Test

`Hypothesis: Lower-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 < p_2 $$
- $p_1$ the probability that a randomly selected obs from the row 1 will be in col 1.

- `P-value`$= P( T \leq T_{Obs})$

- `Decision`: IF p_value < $\alpha$ then REJECT $H_0$


`Hypothesis: Upper-tailed test`
$$ H_0:p_1 = p_2 $$ 
$$H_1: p_1 > p_2 $$
- $p_1$ the probability that a randomly selected obs from the row 1 will be in col 1.

- `P-value`$= P( T \geq T_{Obs})$

- `Decision`: IF p_value < $\alpha$ then REJECT $H_0$

## Fisher's Exact Test - Example 

14 newly hired business majors.
- 10 males and 4 females
- 2 Jobs are needed: 10 Tellers and 4 Account Rep.

`Data` 


|         | Account Rep. | Tellers | `Totals`
|---------|:-----:|:------:|:------:|
| **Males **   |  X=1  |  9   |   r = 10   |
| **Females **   |  3  |  1  |    4   |
| **`Totals`**  | c= 4 |   10 | N = 14  |

 2 x 2 contingency table
 
 `Question`: Test if females are more likely than males to get the account Rep. job.
 
## Fisher's Exact Test - Example
 

1. Define the null and alternative hypotheses <br>
`Answer:` $$ H_0: p_1 \geq p_2 $$ $$H_1: p_1 < p_2 $$
This is an lower-tailed test.
2. Find the Test statistic observed and null distribution<br>
`Answer:` $T_{obs}=X=1$ and $T\sim hypergeometric(14,10,4)$`
3. Find P-value<br>
`Answer:` p-value=`r {phyper(1,10,4,4)}`
5. Decision and Interpretation
<br>
`Answer:` Since P-value < 0.05 then Reject $H_0$.



## Fisher's Exact Test - Example with R

```{r, echo=TRUE}

data = cbind(c(1,3),c(9,1)) # create data
fisher.test(data,alternative = "l")
```

- `Interpretation`: Reject $H_0$. There is evidence to support that the data is compatible with the assumption that females are more likely than males to get the account Rep. job.



## Mantel-Haenszel Test 2x2xk
An extension of Fisher's exact to several 2x2 tables.


```{r}
mydata =array(c(10,12,1,1,9,11,0,1,8,7,0,3),
               dim=c(2,2,3),
               dimnames=list(c("Treat.","Control"),c("Success","Failure"),c("Group 1","Group2","Group 3")))
mydata
mantelhaen.test(mydata,alternative = "g")
```




## The Median Test



Test for equal medians.

$$H_0:  \text{All C populations have the same median} $$
$$H_1:  \text{At least two populations have different medians} $$

`Data`

- C random samples are independent
- Arrange the data as follows:
  - Find the Grand Median (GM), that is the median of the combined samples.
  - Set up a 2 by C contingency table as follows:


|         | Sample 1 | Sample 2 | ...| Sample C|**`Totals`**|
|---------|:-----:|:------:|:------:|:------:|:------:|
| $>$ GM   |  $O_{11}$  |  $O_{12}$ | ... |   $O_{1C}$   |a|
| $\leq$ GM | $O_{21}$  | $O_{22}$ | ...|   $O_{2C}$  |b|
| **`Totals`**  | $n_{1}$ |   $n_{1}$| ...| $n_{C}$  |N|



- `Test Statistic`

$$ T = \frac{N^2}{ab} \sum_{i=1}^{C} \frac{O^2_{1i}}{n_i}  - \frac{Na}{b} $$


Under Null hypothesis: $T \sim \chi^2_{C-1}$; a chi-square distribution with C-1 degrees of freedom.

- `P-value` $=P(T \geq T_{obs})$

- `Decision`: IF p_value < $\alpha$ then REJECT $H_0$




## The Median Test - Example

- 4 methods of growing corn is used.
- The yield per acre is measured and compared across the 4 methods.

`Question`: Do the medians yield per acre differ across the 4 methods.


1. Define the null and alternative hypotheses
$$H_0:  \text{All methods have the same median yield per acre} $$
$$H_1:  \text{At least two of the methods medians differ} $$

2. Set up data: See lecture notes

3. `Test statistic`: 

$$T_{obs} = 17.6$$

Under Null hypothesis: $T \sim \chi^2_{3}$


 
## The Median Test - Example R


```{r}
#| echo: true
  
# install.packages("agricolae")
library(agricolae) # package

data(corn) # data

# The Median Test
median_test_out= Median.test(corn$observation,corn$method)

```


Multiple Comparison 

```{r}
#| echo: true

# Visualization
plot(median_test_out)
```