---
title: "Ranks Based Tests"
format: html
author: Dr. Cohen 
self-contained: true
---

## Wilcoxon rank sum Test or Mann-Whitney Test

-   2 independent samples

### Example 1 : Temperature data

Test that the mean high temperature of city 1 is greater than the mean high temperature of city 2.

```{r}
temp_city1 = c(83,91,89,89,94,96,91,92,90) # X
temp_city2 = c(78,82,81,77,79,81,80,81) # Y
wilcox.test(temp_city1,
            temp_city2,
            alternative = "g",
            exact=FALSE)
```

## Example 2

```{r}
x = c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
y = c(1.15, 0.88, 0.90, 0.74, 1.21)
wilcox.test(x, y, alternative = "g")
```

## Wilcoxon Signed Rank Test

-   2 dependent samples

### Example 1 : Twin data

Test if the first born twin is more aggressive than the second born twin

```{r}
FBT = c(86,71,77,68,91,72,77,91,70,71,88,87) # X
SBT = c(88,77,76,64,96,72,65,90,65,80,81,72) # Y
#in R D= X-Y
wilcox.test(SBT,FBT,
            paired=TRUE,
            alternative="l",
            exact = FALSE,
            conf.int=TRUE)
```

## Example

The cake batter is mixed until it reaches a specified level of consistency. The time required using two mixers A and B.

Find 95 confidence interval of the difference between two means.

```{r}
# Time required 
A=c(7.3,6.9,7.2,7.8,7.2)
B=c(7.4,6.8,6.9,6.7,7.1)

wilcox.test(A,B,alternative = "t",conf.int = TRUE,exact=FALSE)
```

## Kruskal-Walis Test

-   K independent samples
 
### Example 1 - Corn data

```{r}
library(agricolae)
data(corn)
head(corn)
kruskal.test(observation~method,data=corn)
Median.test(corn$observation,corn$method)

## KW test is reject --> post hoc analysis
# install.packages(c("PMCMRplus", "PMCMR"))# install package is needed
library(PMCMRplus)
kwAllPairsConoverTest(corn$observation,corn$method,
                             p.adjust.method = "bon")
kwAllPairsConoverTest(corn$observation,
                             corn$method,p.adjust.method = "none")
```

### Example 2 - Intructors data

3 instructors compared grades they assigned over the past semester:

| Grades | I 1 | I 2 | I 3 |
|--------|:----|:----|:----|
| A      | 4   | 10	  | 6   |
| B      | 14	  | 6	   | 7   |
| C      | 17  | 9   | 8   |
| D      | 6	   | 7   | 6   |
| F      | 2   | 6   | 1   |

Do some instructors tend to give higher grades than other instructors?

The data here is ordinal (A to F). We can code A to be 5 and F=1

```{r}
I1= c(rep(5,4),rep(4,14),rep(3,17),rep(2,6),rep(1,2)) 
I2= c(rep(5,10),rep(4,6),rep(3,9),rep(2,7),rep(1,6))
I3= c(rep(5,6),rep(4,7),rep(3,8),rep(2,6),rep(1,1))

grades = c(I1,I2,I3)
Instr = c(rep("I1",43),rep("I2",38),rep("I3",28))

kruskal.test(grades,Instr)
```

- P-value = 0.8517 is greater than 5% then we fail to reject the null hypothesis. There is evidence to support that the grades assigned by the 3 instructors are not significantly different from each other.


## Friedman Test 

- K dependent samples

### Example 1 - Homeowner data - p. 371

12 homeowners are randomly selected to participate in an experiment with a plant nursery. Each homeowner was asked to select four fairly identical areas in his yard and to plant 4 different types of grasses.

At the end of the experiment, each homeowner was asked to rank the grass types in order of preference. The rank 1 was assigned to the least preferred. 

The hypothesis was that there is no difference in preferences of the grass types. 

```{r}
#| warning: false

library(tidyverse)

hw = (1:12) # homeowner id
grass1 = c(4,4,3,3,4,2,1,2,3.5,4,4,3.5) # preference for grass type 1 
grass2 = c(3,2,1.5,1,2,2,3,4,1,1,2,1)# preference for grass type 2 
grass3 = c(2,3,1.5,2,1,2,2,1,2,3,3,2)# preference for grass type 3 
grass4 = c(1,1,4,4,3,4,4,3,3.5,2,1,3.5)# preference for grass type 4 
df = data.frame(hw,grass1,grass2,grass3,grass4)

df1 = df %>% 
  pivot_longer(cols = -hw, names_to =  "GrassType", values_to = "preference") 

head(df1)

friedman.test(y=df1$preference,df1$GrassType,blocks=df1$hw)

```

Results: Reject the null hypothesis. There is evidence to suggest that preference of grasses types are different among at least two homeowners.

##### Post hoc analysis

```{r}
frdcomp = frdAllPairsConoverTest(y=df1$preference,
                                 df1$GrassType,
                                 blocks=df1$hw,
                                 p.adjust.method = "bon")
frdcomp

boxplot(preference~GrassType, data= df1)
```

Multiple comparison with the Bonferroni correction shows no significance difference. We may conclude the significance results from the Friedman test may be a false positive.


## Correlation coefficient

12 MBA students are studied to measure the strength of the relationship between their GMAT and GPA.

```{r}
gmat=c(710,610,640,580,545,560,610,530,560,540,570,560)
gpa=c(4,4,3.9,3.8,3.7,3.6,3.5,3.5,3.5,3.3,3.2,3.2)

cor.test(gmat,gpa,method = "s",alternative = "t")
cor.test(gmat,gpa,method = "k",alternative = "t")
cor.test(gmat,gpa,method = "p",alternative = "t")
```
